---
title: 'EBI Causality in biomedicine: Day 1 - Fundamental Problem of Causal Inference'
author: "Sjoerd Beentjes and Ava Khamseh"
date: "2026-10-5"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Load auxiliary scripts with code

```{r}
## Source auxiliary scripts: Utilities, data-generation processes (dgp), estimators
source("./scripts/utils.R")
source("./scripts/dgp.R")
source("./scripts/estimators.R")
```

## Load the required packages for these simulations

```{r, echo = TRUE, message = FALSE, warnings = FALSE, results = 'hide'}
# Data manipulation 
library(tidyverse)
library(MASS)
library(survey)
library(boot)

# Plotting and visualisation
library(ggplot2)
library(skimr)

# Learning algorithms
library(glm2)
library(glmnet)
library(randomForest)
library(SuperLearner)
```

# Simulation study 1 (slide 24)

### Data structure $O = (W_1, W_2, A, Y)$

The data is generated from the following data-generation process:

$$
W = (W_1, W_2) \sim \mathcal{N}\left( \begin{pmatrix} 0 \\ 1 \end{pmatrix}, \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix} \right) \\
A \sim \mathcal{B}(g(W)) \, \text{ where } g(W) = \frac{1}{1+e^{-0.5 + 1.5W_1 - 0.5W_2}} \\
Y \sim \mathcal{N}\Bigl(1 + A - 0.7W_1 + 0.3 e^{-W_1W_2}, 1\Bigr)
$$

Note that the true propensity score $e(W) \equiv P(A=1|W)$ is a main-terms logistic regression, and the outcome $Y$ is normally distributed with a mean that depends linearly on treatment $A$, but non-linearly on the covariates $(W_1,W_2)$.

```{r, Generate data}
# Generate data using sim1a() function from "./scripts/dgp.R" + examine data
n_sample <- 2000
data_a <- sim1a(n = n_sample)

# Extract data summary with skimr: Missingness (none here), mean, sd, quantiles, histogram
skim(data_a)

# The true Average Treatment Effect (ATE) is equal to 1
true_value <- 1
```

We specify the linear model $Y \sim A + W_1 + W_2$, i.e., we make the assumption that outcome $Y$ is a linear function of treatment $A$ and covariates $W_1, W_2$. If this is not true, the model is said to be misspecified.

```{r, Model specification}
# Extract the covariates
# For Q, define a baseline linear model Y ~ A + W1 + W2
covariates <- colnames(data)[grepl("^W", colnames(data))]
Q_lin <- as.formula(paste("Y ~ A +", paste(covariates, collapse = " + ")))

# Check the formula is as expected
print(Q_lin)
```

### Obtain the estimate by extracting the coefficient (slide 24)

We fit the linear model to data using Ordinary Least Squares (OLS), extract the coefficient in front of $A$, together with its $95\%$ confidence interval (CI) and corresponding p-value. Crucially, these are computed by using the linear model assumption: If the linear model is misspecified, the CI and p-value are not reliable.

```{r}
# Use Outcome Regression (OR, or OLS) with Q_lin. First, we define the OLS function
OLS_lin = function(data){ estimate_OLS(data, alpha = 0.05, true_value, Qform = Q_lin)}

# Generate data k = 1 times, and tabulate the results (MC_est() from "./scripts/utils.R")
n_sample <- 2000
MC_est(k=1, n=n_sample, simulation=sim1a, est_list=list(OLS_lin = OLS_lin))
```

```{r}
# Repeat the above k = 200 times + save the resulting dataframe
sim1_200 <- MC_est(k=200, n=n_sample, simulation=sim1a, est_list=list(OLS_lin = OLS_lin))
```

```{r}
# Inspect the table of 200 estimates
head(sim1_200)

# Plot the histogram with hist_per_est() from "./scripts/utils.R" (xlimits is optional)
# Note that this step disregards OLS's CI estimate and p-value
sim1_200 %>% hist_per_est(xlimits = c(0, 1.2))

# Compute summary statistics with MC_res() from "./scripts/utils.R"
sim1_200 %>% MC_res()
```

Although coverage is reasonable $\approx 81.5\%$ at this sample size $n = 2000$, the histogram is clearly not centred at the true value of one. It is instructive to examine what happens to the coverage when sample size increases (e.g., set $n_{\text{sample}} = 20000$).

### Work model-independently: Use a substitution estimator (slides 25-30)

The Average Treatment Effect is defined model-independently via the equation

$$
\psi_s = \mathbb{E}_W \bigl[\mathbb{E}(Y | W,A=1) - \mathbb{E}(Y | W,A=0)\bigr].
$$

We can think of this as computing the treatment effect in each covariate stratum given by $W$, and then taking a weighted average to combine all contributions. Next, we estimate the ATE by fitting the Outcome Regression $Q(W,A) = \mathbb{E}(Y|W,A)$ using a linear model and OLS, and then substituting this fit $\hat{Q}_n(W,A)$ into the above formula. This yields the estimator

$$
\hat{\psi}_n^{\text{OR}} = \frac{1}{n} \sum_{i=1}^n \Bigl[\hat{Q}_n(W_i,1) - \hat{Q}_n(W_i,0)\Bigr].
$$

In other words, for each data point $O_i$ with treatment $A_i = a$ we predict the expected outcome under treatment given their covariate value $W_i$, namely $\hat{Q}_n(W_i,1)$, as well as the expected outcome under control given their covariate value $W_i$, namely $\hat{Q}_n(W_i, 0)$. In the substitution estimator, these two predictions are contrasted and then averaged over all data points.

```{r}
# Explicitly evaluate the OR substitution estimator
data <- sim1a(n_sample)

# Obtain $\hat{Q}_n(W,A)$ by fitting Q_lin to data
Qfit <- glm2(Q_lin, data = data)
  
# From the fitted model, predict 
covariates <- colnames(data)[grepl("^W", colnames(data))]
Qfit_0W <- predict(Qfit, newdata = data.frame(A = 0, data[covariates]))
Qfit_1W <- predict(Qfit, newdata = data.frame(A = 1, data[covariates]))

# Evaluate estimate
mean(Qfit_1W - Qfit_0W)
```

To obtain a CI and p-value on this substitution estimator, we can use the bootstrap. However, this is relatively slow because it involves re-sampling the data R = 1000 times, each time fitting the substitution estimator. This functionality can be found in the function estimate_OLS_sub in "./scripts/estimators.R"

```{r}
# Define the OR substitution estimator as a function of data
OLS_sub_lin <- function(data){ estimate_OLS_sub(data, alpha=0.05, true_value, Qform=Q_lin)}

# Generate data k = 1 times, and tabulate the results (MC_est() from "./scripts/utils.R")
MC_est(k=1, n=n_sample, simulation=sim1a, est_list=list(OLS_sub = OLS_sub_lin))
```

The above can again be repeated $k = 200$ times to plot the distribution of the central estimates. Again, this plot disregards the error bars, CI, and p-value computed for each individual run (here, via bootstrap).

NOTE: This step is a little time consuming due to the bootstrap.

```{r}
# Repeat the above k = 200 times + save the resulting dataframe
sim1_sub_200 <- MC_est(k=200, n=n_sample, simulation=sim1a, est_list=list(OLS_lin = OLS_lin))
```

```{r}
# Inspect the table of 200 estimates
head(sim1_sub_200)

# Plot the histogram with hist_per_est() from "./scripts/utils.R" (xlimits is optional)
# Note that this step disregards OLS's CI estimate and p-value
sim1_sub_200 %>% hist_per_est(xlimits = c(0, 1.2))

# Compute summary statistics with MC_res() from "./scripts/utils.R"
sim1_sub_200 %>% MC_res()
```

### Compare a linear model vs a model with interaction terms (slide 31)

We specify the alternative model $Y \sim A + W_1 + W_2 + AW_1 + AW_2$ and investigate both OR estimators. For the coefficient estimator, we again extract the coefficient in front of $A$ although other choices can be argued; this is the problem! In contrast, the OR substitution estimator is defined model-independently, so its definition does not depend on the model/algorithm chosen to fit the data.

```{r}
# Define the interaction model
Q_int <- as.formula(paste("Y ~ W1:W2 +", "A*(", paste(covariates, collapse = " + "), ")"))
Q_base <- "Y ~ A"

# List the four OR estimators: Coefficient/substitution and linear/interaction
est_OLS_vs <- list(
  OLS = function(data){ estimate_OLS(data, Qform = Q_lin)},
  OLS_int = function(data){ estimate_OLS(data, Qform = Q_int)},
  OLS_sub = function(data){ estimate_OLS_sub(data, Qform = Q_lin)},
  OLS_sub_int = function(data){ estimate_OLS_sub(data, Qform = Q_int)}
)
```

Next, we fit all four models on $k= 200$ draws of the data. This is somewhat time consuming because there are two algorithms (the two substitution models) which combine CI and p-value via bootstrap.

```{r}
## Perform MC experiment
# Smallest sample
n_sample = 2000
sim1_OLS_vs_2K <- MC_est(k = 200, n_sample, simulation = sim1a, est_OLS_vs)
```

Visualise and plot:

```{r}
# Histogram the central estimates, per estimator
sim1_OLS_vs_2K %>% hist_per_est(xlimits = c(0,1.5))

# Plot all four histograms together
sim1_OLS_vs_2K %>% hist_all(xlimits = c(0,1.5))

# Histogram the lower and upper 95% CI bounds, per estimator
sim1_OLS_vs_2K %>% hist_per_est_CI(xlimits = c(0, 1.5))

# Display summary statistics, such as bias, variance, and bias.to.se ratio
sim1_OLS_vs_2K %>% MC_res()
```

Re-doing the above analysis for larger sample sizes (e.g., $n_{\text{sample}} = 20,\!000$ on slide 32) shows that for all four estimators coverage decreases to zero, and the bias.to.se ratio becomes large ($\gg 1$). This indicates that the bias is so large that it exceeds the estimation uncertainty as quantified by the standard error (se).

## Building up the TMLE estimator from scratch

In this section, we construct the TMLE estimator step-by-step. TMLE estimators can be given additional robustness properties. The function estimator_tmle() in the file "./scripts/estimators.R" has some of these.

### Step 1: Fit nuisance functions $Q(W,A)$ and $g(W)$

For the outcome model, we (correctly) specify the linear model with interaction term$$
Y \sim A + W_1 + W_2 + AW_1 + AW_2 + W_1W_2.
$$Thus, we expect the OR substitution estimator to perform well in the sense that generating $k$ data sets and applying the OR substitution estimator $k$ times should result in a histogram centered around the true value.

For the propensity score $g(W) = P(A=1|W)$, we (incorrectly) specify a main-term logistic model $\operatorname{logit} g(W) \sim W_1 + W_2$. The Inverse Probability of Treatment Weighting (IPTW) will likely not perform well.

```{r}
# Generate data from sim1a()
n_sample <- 2000
data <- sim1b(n_sample)

# Define the outcome model (Q) and propensity score model (g)
covariates <- colnames(data)[grepl("^W", colnames(data))]
Q_int2 <- as.formula("Y ~ A + W1 + W2 + A*W1 + A*W2 + W1*W2")
g_lin <- as.formula("A ~ W1 + W2")

# Outcome mechanism: Fit Q to the data
# Record predicted values under treatment (A=1) and control (A=0) for all n_sample data points
# Qform <- "Y ~ A + W1 + W2"
Q_init <- glm(Q_int2, family = gaussian, data = data)
data$Q0W <- predict(Q_init, newdata = data.frame(A = 0, data[covariates]))
data$Q1W <- predict(Q_init, newdata = data.frame(A = 1, data[covariates]))
data$Q <- predict(Q_init, type = "response")

# Treatment mechanism: Fit g to the data
# Record  predicted treatment probabilities g(W) = P(A=1|W) for all n_sample data points
gform <- "A ~ W1+W2"  #g_lin
g_init <- glm(gform, family = binomial, data = data)
data$g1W <- predict(g_init, type = "response")
```

At this point, we can evaluate the OR substitution estimator. This gives our initial ATE estimate:

```{r}
# The initial ATE estimate (without guarantees or error bars)
ATE_init <- mean(data$Q1W - data$Q0W)
ATE_init
```

Note that this value is close to the true value of 1 since the outcome model is correctly specified. Selecting a misspecified model for Qform (e.g., a basic model $Y \sim A$ ) will lead to very different results. (Try!)

### Step 2: Update the initial estimate via the TMLE step

Our initial fit $\hat{Q}_n(W,A)$ can be updated using the TMLE step, via an additional linear regression:

$$
\hat{Q}_{n,\epsilon} = \hat{Q}_n(A,W) + \epsilon H(\hat{g}_n)(W,A),
$$

where the parameter $\epsilon$ is fitted using Maximum Likelihood Estimation (MLE). The expression multiplying $\epsilon$ is known as the *clever covariate*, which is specific to the target quantity of interest. For us, the quantity of interest is the ATE, and the clever covariate for ATE can be shown to equal

$$
H(\hat{g}_n)(W,A) = \frac{A}{\hat{g}_n(W)} - \frac{1-A}{1-\hat{g}_n(W)}.
$$

This explains the name *Targeted Maximum Likelihood Estimation* or *TMLE*.

First, we compute the clever covariate:

```{r}
# We use the propensity score fit g_init obtained above
data$H <- (data$A / data$g1W) - ((1 - data$A) / (1 - data$g1W))
```

Second, we perform a linear regression of outcome $Y$ on the clever covariate to obtain $\epsilon$ via MLE whilst including the initial fit $\hat{Q}_n(W,A)$ as offset:

```{r}
# Fit \epsilon via linear regression (MLE):
TMLE_fit <- glm2(data$Y ~ -1 + H + offset(data$Q), family = gaussian, data = data)
epsilon_fit <- coef(TMLE_fit)

# Check how large epsilon is
epsilon_fit
```

Third, we update our initial $Q$ fit by setting

$$
\hat{Q}^*_n(W,A) = \hat{Q}_n(W,A) + \hat{\epsilon}_n H(\hat{g}_n)(W,A).
$$

It is helpful to unpack this equation for treatment ($A=1$) and control ($A=0$) separately:

\begin{align}
\hat{Q}^*_n(W,1) &= \hat{Q}_n(W,1) + \frac{\hat{\epsilon}}{\hat{g}_n(W)} \\
\hat{Q}^*_n(W,0) &= \hat{Q}_n(W,0) - \frac{\hat{\epsilon}}{1-\hat{g}_n(W)}
\end{align}

In particular, individual prediction by the initial fits are shifted in opposite directions for treatment and for control. The updating step corrects bias in the initial Q fit by leveraging the propensity score fit $\hat{g}_n(W)$. The worse the initial Q fit (e.g., try using $Y \sim A$), the larger the residual bias due to misspecification of the $Q$-model, the larger the correction due to the TMLE step.

```{r}
# Update initial fit via the TMLE step
data$Q_star1 <- data$Q1W + epsilon_fit * (1 / data$g1W)
data$Q_star0 <- data$Q0W - epsilon_fit * (1 / (1 - data$g1W))

# Final TMLE estimate
ATE_tmle <- mean(data$Q_star1 - data$Q_star0)
ATE_tmle
```

It is well known that (i) fitting $\epsilon$ with a *weighted* regression, and (ii) scaling $Y$ to lie in the unit interval $[0,1]$ and then using *logistic regression* for the TMLE update step, yields better performance. These are some of the additional robustness properties mentioned above with which TML estimators can be equipped.

```{r}
# Scale outcome to [0,1]
a <- min(data$Y)
b <- max(data$Y)
data$Y <- (data$Y - a) / (b - a)

# Re-fit Q now that Y is re-scaled
Qform <- "Y ~ A + W1 + W2"
Q_init <- glm(Qform, family = gaussian, data = data)
data$Q0W <- predict(Q_init, newdata = data.frame(A = 0, data[covariates]))
data$Q1W <- predict(Q_init, newdata = data.frame(A = 1, data[covariates]))
data$Q <- predict(Q_init, type = "response")
```

```{r}
# Define weights, and weighted clever covariate
data$weights <- ifelse(data$A, 1/data$g1W, 1/(1-data$g1W))
data$H_w <- data$A - (1 - data$A)

# Fit \epsilon_w via weighted, logistic regression
wTMLE_fit <- glm2(data$Y ~ -1 + H_w + offset(logit(data$Q)),
                  family = quasibinomial, data = data, weights = weights)
epsilon_w <- coef(wTMLE_fit)
epsilon_w

# Update the initial Q-fits
data$wQ_star1 <- expit(logit(data$Q1W) + epsilon_w)
data$wQ_star0 <- expit(logit(data$Q0W) - epsilon_w)

# Evaluate the weighted logistic TMLE
ATE_wtmle <- mean(data$wQ_star1 - data$wQ_star0)*(b-a)
ATE_wtmle
```

The final TMLE estimator is similar to the OR substitution estimator, but using the updated $Q$-fits:

\begin{align}
\hat{\psi}_n^{\text{tmle}} &= \frac{1}{n} \sum_{i=1}^n \Bigl[\hat{Q}^*_n(W_i,1) - \hat{Q}^*_n(W_i,0)\Bigr] \\
&= \frac{1}{n} \sum_{i=1}^n \left[ \left\{\hat{Q}_n(W_i,1) +\frac{\hat{\epsilon}}{\hat{g}_n(W_i)}\right\} - \left\{\hat{Q}_n(W_i,0) - \frac{\hat{\epsilon}}{1-\hat{g}_n(W_i)}\right\}\right]
\end{align}

It is helpful to visualise the change induced by the TMLE step.

```{r}
# Another try
vis_TMLE2 <- data %>%
  dplyr::select(c(g1W,Q0W,Q1W,Q_star0,Q_star1,wQ_star0,wQ_star1)) %>% 
  slice_sample(n = 20) %>% 
  mutate(x_position = seq_len(n()))

# Plot Q0W and Q1W vs wQ_star0 and wQ_star1
ggplot(vis_TMLE2, aes(x = x_position)) +
  geom_point(aes(y = Q0W), color = "grey") +
  geom_point(aes(y = Q1W), color = "grey") +
  geom_point(aes(y = wQ_star0), color = "blue") +
  geom_point(aes(y = wQ_star1), color = "blue") +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  theme_minimal() +
  labs(x = "Index", y = "Values")

# Plot Q0W and Q1W vs Q_star0 and Q_star1 vs g1W
ggplot(vis_TMLE2, aes(x = x_position)) +
  #geom_point(aes(y = g1W), color = "black") +
  #geom_point(aes(y = 1-g1W), color = "black") +
  geom_point(aes(y = Q0W), color = "grey") +
  geom_point(aes(y = Q1W), color = "grey") +
  geom_point(aes(y = Q_star0), color = "red") +
  geom_point(aes(y = Q_star1), color = "red") +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  theme_minimal() +
  labs(x = "Index", y = "Values")

# All together
ggplot(vis_TMLE2, aes(x = x_position)) +
  geom_point(aes(y = Q0W), color = "grey") +
  geom_point(aes(y = Q1W), color = "grey") +
  geom_point(aes(y = wQ_star0), color = "blue") +
  geom_point(aes(y = wQ_star1), color = "blue") +
  geom_point(aes(y = Q_star0), color = "red") +
  geom_point(aes(y = Q_star1), color = "red") +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  theme_minimal() +
  labs(x = "Index", y = "Values")

```

### Step 3: Estimate the variance, CI, and p-value

TMLE theory shows that, provided the $\hat{Q}_n$ and $\hat{g}_n$ are fitted well enough (this can be quantified), then the TMLE estimator $\hat{\psi}_n^{\text{tmle}}$ is normally distributed for large enough sample size, namely

$$
\sqrt{n}\bigl(\hat{\psi}_n^{\text{tmle}} - \psi_s \bigr) = \frac{1}{\sqrt{n}}\sum_{i=1}^n D_{P_0}(O_i) + o_P(1) \rightsquigarrow \mathcal{N}\bigl(0,\text{Var}D_{P_0}\bigr).
$$

Here $D_{P_0}(O_i) = D(Q_0,g_0)(O_i)$ is the *influence function* (IF) of the statistical estimand $\psi_s$. One can think of this as a first order derivative. Each target parameter, such as the ATE, has its own IF (if it exists). The IF is a function of the data $O_i$, and depends on the $Q$ and $g$ fits. For ATE, the IF is

$$
D(Q,g)(O) = H(g)(W,A) \bigl\{Y - Q(W,A) \bigr\} + Q(W,1) - Q(W,0) - \psi_s.
$$

Here we recognise the clever covariate $H(g)(W,A) = \frac{A}{g(W)} - \frac{1-A}{1-g(W)}$.

For many parameters of interested, e.g., ATE, ATT, survival time, optimal treatment rule, their influence functions have been computed, or can be computed using a standard toolbox.

VIa the theoretical result above, the variance of the IF gives us access to the variance of the TMLE estimator.

```{r}
# Estimate the variance of the IF
data$IF <- data$A * data$H * (data$Y - data$Q_star1) - (1-data$A) * data$H * (data$Y - data$Q_star0) + (data$Q_star1 - data$Q_star0) - ATE_init

# Estimate the sample variance and standard error of the IF
se <- sd(data$IF)
se
```

Note that the standard deviation of the IF is *NOT* the standard error on the TMLE estimate. To obtain the standard error of the TMLE estimator, we first need to divide by the square root of the sample size.

Via the asymptotic normal distribution, an asymptotic $95\%$ confidence interval of the TMLE estimator is $$
\Bigl(\hat{\psi}_n^{\text{tmle}} \pm 1.96 \sqrt{\operatorname{Var}(IF)/n} \Bigr).
$$

```{r}
# Confidence interval
CI_lower <- ATE_tmle - 1.96*se / sqrt(n_sample)
CI_upper <- ATE_tmle + 1.96*se / sqrt(n_sample)

# Print CI
data.frame(Est.initial = ATE_init, Estimate = ATE_tmle, CI_lower = CI_lower, CI_upper = CI_upper)
```

The CI contains the true value, as expected from theory. In fact, a 95% CI is expected to contain the true value 95% of the times (so not always!), and the actual performance is quantified by coverage estimates.

# Simulation study 2 (slide 43)

### Data structure $O = (W_1, W_2, A, Y)$

For the second simulation (slide 43), we modify the data-generation process as follows:

$$
W = (W_1, W_2) \sim \mathcal{N}\left( \begin{pmatrix} 0 \\ 1 \end{pmatrix}, \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix} \right) \\
A \sim \mathcal{B}(e(W)) \, \text{ where } e(W) = \frac{1}{1+e^{-0.5 + 1.5W_1W_2}} \\
Y \sim \mathcal{N}\Bigl(1 + A - 0.5W_1 - 0.25 W_1W_2, 1\Bigr)
$$

Note that the true propensity score $e(W) \equiv P(A=1|W)$ is now a logistic regression with interaction term $W_1W_2$, and the outcome $Y$ is normally distributed with a mean that depends on the covariates $(W_1,W_2)$.

```{r, Generate data}
# Generate data using sim1b() function from "./scripts/dgp.R" + examine data
n_sample <- 2000
data_b <- sim1b(n = n_sample)

# Extract data summary with skimr
skim(data_b)

# The true Average Treatment Effect (ATE) is equal to 1
true_value <- 1
```

### Further robustifying TMLE on simulation 2

There are three standard ways to robustify the TMLE estimation procedure:

1.  Using two coefficients $(\epsilon_1,\epsilon_0)$ to fit the updating step for $Q(1,W)$ and $Q(0,W)$

2.  Fitting these $(\epsilon_1, \epsilon_0)$ with a *weighted* regression for stability, and

3.  Scaling outcome $Y$ to lie in the unit interval $[0,1]$, using *logistic regression* for the TMLE update step so that predicted outcome still lies in the unit interval $[0,1]$, and then re-scaling $Y$ back to its original units. This procedure ensures natural bounds on outcome $Y$ are respected during the TMLE step.

Below, we see an example where a basic TMLE is used ignoring 1-3 above. This is followed by an implementation of TMLE that uses all three robustification steps, which has better performance.

```{r}
# Generate data
data <- sim1b(n_sample)

# Define the outcome model (Q) and propensity score model (g)
covariates <- colnames(data)[grepl("^W", colnames(data))]
Q_int2 <- as.formula("Y ~ A + W1 + W2 + A*W1 + A*W2 + W1*W2")
g_lin <- as.formula("A ~ W1 + W2")

# Outcome mechanism: Fit Q to the data
# Record predicted values under treatment (A=1) and control (A=0) for all n_sample data points
Qform <- "Y ~ A + W1 + W2"
Q_init <- glm(Qform, family = gaussian, data = data)
data$Q0W <- predict(Q_init, newdata = data.frame(A = 0, data[covariates]))
data$Q1W <- predict(Q_init, newdata = data.frame(A = 1, data[covariates]))
data$Q <- predict(Q_init, type = "response")

# Treatment mechanism: Fit g to the data
# Record  predicted treatment probabilities g(W) = P(A=1|W) for all n_sample data points
gform <- "A ~ W1+W2"  #g_lin
g_init <- glm(gform, family = binomial, data = data)
data$g1W <- predict(g_init, type = "response")

# The initial ATE estimate (without guarantees or error bars)
ATE_init <- mean(data$Q1W - data$Q0W)
ATE_init

# We use the propensity score fit g_init obtained above
data$H <- (data$A / data$g1W) - ((1 - data$A) / (1 - data$g1W))

# Fit \epsilon via linear regression (MLE):
TMLE_fit <- glm2(data$Y ~ -1 + H + offset(data$Q), family = gaussian, data = data)
epsilon_fit <- coef(TMLE_fit)

# Check how large epsilon is
epsilon_fit

# Update initial fit via the TMLE step
data$Q_star1 <- data$Q1W + epsilon_fit * (1 / data$g1W)
data$Q_star0 <- data$Q0W - epsilon_fit * (1 / (1 - data$g1W))

# Final TMLE estimate
ATE_tmle <- mean(data$Q_star1 - data$Q_star0)
ATE_tmle

# TMLE lower and upper 95% confidence intervals
data$IF <- data$A * data$H * (data$Y - data$Q_star1) - (1-data$A) * data$H * (data$Y - data$Q_star0) + (data$Q_star1 - data$Q_star0) - ATE_tmle
ATE.ci.lower <- ATE_tmle - 1.96*sqrt(var(data$IF) / n_sample)
ATE.ci.upper <- ATE_tmle + 1.96*sqrt(var(data$IF) / n_sample)

# Print
ATE.ci.lower
ATE.ci.upper
```

Compare with the constructed robustified TMLE function of PhD student Asantewaa Sarpong. This function, called estimate_TMLE(), can be found in "./scripts/estimators.R". Here we equip it with the correctly specified outcome regression Q_int2, and the incorrectly specified main-term logistic regression g_lin.

```{r}
# Generate + fit with TMLE (100 times)
estimate_TMLE_AS <- function(data){ estimate_TMLE(data, Qform = Q_int2, gform = g_lin)}
#res_AS <- MC_est(k = 100, n_sample = 1000, sim1b, est_list = list(TMLE = estimate_TMLE_AS))
```

```{r}
# Produce results table
res_AS %>% MC_res()

# Plot histogram
res_AS %>% hist_all(xlimits = c(0.5,1.5))
```

## Slides 44-45: Comparison of TMLE with OR substitution and IPTW

Using the same Simulation study 2 as above, we correctly specify the outcome regression Q_int2 as

$$
Y \sim A + W_1 + W_2 + A W_1 + A W_2,
$$

and *incorrectly* specify the propensity score as main-terms logistic regression $\operatorname{logit} g(W) \sim W_1 + W_2$. We expect the OR substitution estimator to perform well, IPW poorly, and TMLE well.

```{r}
# Data
data <-sim1b()

# Define the outcome model (Q) and propensity score model (g)
covariates <- colnames(data)[grepl("^W", colnames(data))]
Q_int2 <- as.formula("Y ~ A + W1 + W2 + A*W1 + A*W2 + W1*W2")
g_lin <- as.formula("A ~ W1 + W2")

# Define the list of estimators: OR with Q_int2, IPTW with g_lin, TMLE with Q_in2 and g_lin
ests_sim1b <- list(
  OLS_Qint2 = function(data){ estimate_OLS(data, Qform = Q_int2)},
  IPW_glin = function(data){ estimate_IPW(data, gform = g_lin)},
  TMLE_Qint2_glin = function(data){ estimate_TMLE(data, Qform = Q_int2, gform = g_lin)}
)
```

```{r}
# Perform the estimation 200 times at sample sizes (i) n = 2000, and (ii) n = 20,000
sim1b_2K <- MC_est(k = 200, n_sample = 2000, simulation = sim1b, ests_sim1b)
sim1b_20K <- MC_est(k = 200, n_sample = 20000, simulation = sim1b, ests_sim1b)
```

Visualise and plot the results at sample size n = 2000. TMLE perform best across various metrics, i.e., nominal coverage of 95% (not overly conservative, not anti-conservative, but as expected), least average bias, smallest mean variance, and smallest bias.to.se ratio indicating an optimal bias-variance trade-off. This display good finite (smaller) sample performance.

```{r}
# Histogram the central estimates, per estimator
sim1b_2K %>% hist_per_est(xlimits = c(0,1.5))

# Plot all four histograms together
sim1b_2K %>% hist_all(xlimits = c(0,1.5))

# Histogram the lower and upper 95% CI bounds, per estimator
sim1b_2K %>% hist_per_est_CI(xlimits = c(0, 1.5))

# Display summary statistics, such as bias, variance, and bias.to.se ratio
sim1b_2K %>% MC_res()
```

Visualise and plot the results at sample size n = 20,000. TMLE and correctly specified OR substitution estimator perform comparably, with OR substitution slightly better in various metrics at this sample size.

```{r}
# Histogram the central estimates, per estimator
sim1b_20K %>% hist_per_est(xlimits = c(0,1.5))

# Plot all four histograms together
sim1b_20K %>% hist_all(xlimits = c(0,1.5))

# Histogram the lower and upper 95% CI bounds, per estimator
sim1b_20K %>% hist_per_est_CI(xlimits = c(0, 1.5))

# Display summary statistics, such as bias, variance, and bias.to.se ratio
sim1b_20K %>% MC_res()
```

## Slide 46: TMLE corrects initial misspecification in Q with correct g

Using the same Simulation 2, we correctly specify the propensity score $\operatorname{logit} g(W) \sim W_1 \cdot W_2$ but incorrectly specify the outcome regression $Q(W,A)$ in three different ways. Despite the misspecification of Q, TMLE will still estimate the causal effect correctly. This illustrates its **double robust** property.

We select the following three misspecified outcome models:

\begin{align}
Q_{\text{con}} \, &\colon Y \sim 1 \\
Q_{\text{base}} \, &\colon Y \sim A \\
Q_{\text{sin}} \, &\colon Y \sim A + \sin(A \cdot W_1).
\end{align}

The first model fits outcome $Y$ as a constant (ignoring any influence of treatment $A$), the second model only takes treatment $A$ into account but ignores confounders, whereas the third model posits that $Y$ depends on treatment $A$ linearly as well as via a sinusoidal term $\sin(A \cdot W_1)$. Needless to say, all three models are misspecified to some extent as they do not reflect how outcome is generate in Simulation 2.

```{r}
# Define three TMLE estimators with the above OR models, and correct g model
Q_con <- as.formula("Y ~ 1")
Q_base <- as.formula("Y ~ A")
Q_sin <- as.formula("Y ~ A + sin(A*W1)")
g_int2 <- as.formula("A ~ W1:W2")

# Estimators
ests_DR_tmle <- list(
  TMLE_Qcon = function(data){ estimate_TMLE(data, Qform = Q_con, gform = g_int2)},
  TMLE_Qbase = function(data){ estimate_TMLE(data, Qform = Q_base, gform = g_int2)},
  TMLE_Qsin = function(data){ estimate_TMLE(data, Qform = Q_sin, gform = g_int2)}
)
```

Perform the estimation:

```{r}
# Estimate k = 1 time at sample size n = 1000
res_TMLE_DR_1K <- MC_est(k = 1, n_sample = 1000, simulation = sim1b, ests_DR_tmle)
```

Visualise and plot, showing that the TMLE estimator corrects the initial biased estimates due to the misspecified OR models via its TMLE update step with the correct specified propensity model:

```{r}
# Plot the initial estimate (in red) + the TMLE estimate with CI (in black)
res_TMLE_DR_1K %>% MC_res()
res_TMLE_DR_1K[1:3,] %>% plot_est_CI(xlimits = c(0,2)) + 
  geom_point(aes(x = Algorithm, y = Initial.Est), color = "red", size = 3)

#res_TMLE_DR_1K %>% plot_est_CI(xlimits = c(-1,5))
```

# Slides 47-48: Simulation study 3

### Data structure $O = (W_1, W_2, W_3, W_4, W_5, W_6, W_7, W_8, A, Y)$

For the third simulation (slides 47-48), we modify the data-generation process as follows:

\begin{align}
W_1, W_2, W_3 &\sim \mathcal{B}(0.5) \\
W_4 \mid W_1 &\sim \mathcal{B}(0.2+0.5W_1) \\
W_5 \mid W_1, W_2,W_3,W_4 &\sim \mathcal{B}(0.05+0.3W_1+0.1W_2+0.05W_3+0.4W_4) \\
W_6 \mid W_5 &\sim \mathcal{B}(0.2+0.6W_5) \\
W_7 \mid W_3 &\sim \mathcal{B}(0.5+0.2W_3) \\
W_8 \mid W_2, W_6, W_7 &\sim \mathcal{B}(0.1+0.2W_2+0.3W_6+0.1W_7) \\
\operatorname{logit} g(W) &\sim -0.05+0.1W_1+0.2W_2+0.2W_3-0.02W_4-0.6W_5-0.2W_6-0.1W_7 \\
Y &\sim \mathcal{N}\bigl(10+A+W_1+W_2+W_4+2W_6+W_7, 1\bigr)
\end{align} Note that the true propensity score $e(W) \equiv P(A=1|W)$ is a main-term logistic regression which depends on all covariates apart from $W_8$. The outcome $Y$ is normally distributed with a mean that depends linearly on the treatment as well as five of the eight covariates, namely $(W_1,W_2,W_4,W_6,W_7)$.

```{r, Generate data}
# Generate data using sim2() function from "./scripts/dgp.R" + examine data
n_sample <- 2000
data_b <- sim2(n = n_sample)

# Extract data summary with skimr
skim(data_b)

# The true Average Treatment Effect (ATE) is equal to 1
true_value <- 1
```

In this simulation, we show that TMLE with a correctly specified propensity score can be (i) more precise than IPW, and (ii) correct OLS' loss of power when using an overly large regression (i.e. too many terms). This is a cautionary tale that "adding all covariates and interactions just to be sure" in a standard multivariate regression approach can dilute signal; a targeted approach can recover this signal.

```{r}
# Extract the covariates variables
data2 <- sim2()
covariates <- colnames(data2)[grepl("^W", colnames(data2))]

# Declare the regression and propensity score formulas
Q_lin <- as.formula(paste("Y ~ A +", paste(covariates, collapse = " + ")))
Q_int <- as.formula(paste("Y ~", "A*(", paste(covariates, collapse = " + "), ")"))
g_lin <- as.formula(paste("A ~", paste(covariates, collapse = " + ")))
```

Note that $Q_{\text{lin}}$ is the linear model $Y \sim A + W_1 + W_2 + \ldots + W_7 + W_8$ containing all eight covariates, including the three covariates $W_3,W_5,W_8$ that do not appear in the true model. The interaction model $Q_{\text{int}}$ additionally contains all interaction terms between treatment $A$ and covariates $W_1, \ldots, W_8$

$$
Q_{\text{int}} \, \colon \, Y \sim A + (W_1 + \ldots + W_8) + (AW_1 + \ldots + AW_8).
$$

This models is more complex than necessary relative to the true outcome model. Finally, $g_{\text{lin}}$ is a standard main-term logistic regression which is (almost) correctly specified, as it contains the extra covariate $W_8$.

Next, we select our estimators, which consist of

-   OLS_Qtrue: OLS with the oracle (ground truth) model $Y \sim A+W_1+W_2+W_4+W_6+W_7$

-   OLS_Qint: OLS with the (overly complex) model with interactions $Q_{\text{int}}$

-   IPW_glin: IPW with the (almost correctly specified) main-terms logistic regression $g_{\text{lin}}$

-   TMLE_Qint_glin: TMLE with the (overly complex) $Q_{\text{int}}$ model and (almost correct) $g_{\text{lin}}$ model.

```{r}
# Define estimators
ests_OLS_tmle <- list(
  OLS_Qtrue = function(data){ estimate_OLS(data, Qform = "Y~A+W1+W2+W4+W6+W7")},
  OLS_Qint = function(data){ estimate_OLS(data, Qform = Q_int)},
  IPW_glin = function(data){ estimate_IPW(data, gform = g_lin)},
  TMLE_Qint_glin = function(data){ estimate_TMLE(data, Qform = Q_int, gform = g_lin)}
)

# Generate data once, apply all four estimators, and plot the estimates + 95% CI 
MC_est(k = 1, 2000, sim2, ests_OLS_tmle) %>% plot_est_CI()
```

```{r}
# Repeat the same 200 times, and generate summary statistics to assess overall performance
res_OLS_TMLE_2K <- MC_est(k = 100, n_sample = 2000, simulation = sim2, ests_OLS_tmle)

# Investigate performance
res_OLS_TMLE_2K %>% MC_res()
```

We see that TMLE performs as well as the OLS with the oracle (true unknowable) regression model, in terms of bias, mean variance, sample variance, and coverage, with variance \~6 times smaller.
